Can Exposure To Forgiving AI Foster Cooperative Play?

### 2024-06-26 article.Rmd
trstee HMM fit & joint state analytics

# Find IDs for those who participated in experiment and add RS_score and high_RS as new columsn to d_finished.
# Merging d_finished with relevant columns from RS_data based on matching ids
d_finished <- d_finished %>%
  inner_join(RS_data %>% dplyr::select(id, RS_score, high_RS), by = "id")

save(d_finished, file = "data/d_finished.RData")

@@

# View the summarized counts
print(temp)
```

@@

d_finished <- d_finished  %>% mutate( gameNum.f = factor(gameNumber,labels = c("pre", "expo1", "expo2", "expo3","post"),
                                      levels=c("1", "11", "12", "13", "2")),
                                      phase.f = factor(ifelse(gameNumber %in% c("1", "2"), 
                                                       ifelse(gameNumber == "1", "pre", "post"), NA), levels = c("pre", "post"))) %>% 
                              rename(# Before Exposure
                            rating_cooperative_1 = `ratePLayer_trustGame_1.How cooperative do you think the other player is?`,
                            rating_forgiving_1 = `ratePLayer_trustGame_1.How forgiving do you think the other player is?`,
                            rating_playAgain_1 = `ratePLayer_trustGame_1.If given the choice, would you like to play again with the other player?`,

@@

                            )
```

@@

# Calculate means using vectorized operations
d_finished <- d_finished %>%
  mutate(
    LPFS_score = rowMeans(dplyr::select(., all_of(lpfs_cols)), na.rm = TRUE),
    LPFS_self_score = rowMeans(dplyr::select(., all_of(lpfs_1_to_6_cols)), na.rm = TRUE),
    LPFS_other_score = rowMeans(dplyr::select(., all_of(lpfs_7_to_12_cols)), na.rm = TRUE)
  )

@@

avg_ret_df <- d_finished %>%
  dplyr::select(playerId,roundType,investment,returns,roundNum,gameNum.f,condition.f, phase.f, high_RS, LPFS_score, RS_score) %>% 
  filter(roundType=="trust",!is.na(roundNum)) %>% 
  mutate(roundNum = as.numeric(as.character(roundNum))) %>%
  mutate(inv_scaled = as.vector(scale(investment))) %>% 
  mutate(inv_pct = investment/20, ret_pct_0 = ifelse(investment==0,0,returns/(3*investment)),ret_pct_na = ifelse(investment==0,NA,returns/(3*investment)))

@@

```{r modAllReturns, include=FALSE, cache=TRUE}

prepost_df <- avg_ret_df %>% filter(gameNum.f %in% c("pre","post"))

mod_returns_pct_RS <- mixed( ret_pct_na ~ phase.f*condition.f*inv_scaled*high_RS + (1 + phase.f| playerId), prepost_df, REML= TRUE, method="KR")

@@

```{r}
contrasts(prepost_df$phase.f)
contrasts(prepost_df$condition.f)
contrasts(prepost_df$high_RS)

```
```{r}
#RS_model <- lm( RS_score ~ ret_pct_na*inv_scaled + (1 + phase.f| playerId), prepost_df, REML= TRUE, method="KR")
```

@@

# Compute EMMs for the interaction of interest
emms <- emmeans(mod_returns_pct_RS, ~ phase.f * condition.f | high_RS, pbkrtest.limit = 6300)

@@

@@

# Generate the plot with points and error bars
ret_all <- afex_plot(mod_returns_pct_RS, x = "condition.f", trace = "phase.f", dodge = 0.8, error = "within",
                     mapping = c("linetype", "shape", "fill"),
                     factor_levels = list(condition.f = c("Control", "Manipulation")),
                     legend_title = "Phase")

@@

# Convert dataset from wide to long format, focusing on columns that start with "State_"
d_long <- d_finished %>%
  gather(key = "state_col", value = "investorState", starts_with("State_")) %>%
  mutate(
    round_extracted = as.numeric(str_extract(state_col, "(?<=State_r)\\d+")),
    game_extracted = as.numeric(str_extract(state_col, "(?<=game_)\\d+"))
  ) %>%
  # Filter to keep rows where extracted round/game numbers match the actual round/game numbers
  filter(round_extracted == roundNum, game_extracted == gameNumber) %>%
  # Select and rename the desired columns
  dplyr::select(id, condition.f, gameNum.f, gameNumber, roundNum, investorState)

# View the resulting dataset
print(d_long)

@@

# Filter for only games 1 and 2
filtered_data <- d_long %>%
  filter(gameNum.f %in% c("pre", "post"))

@@

# Create the histogram with stacked bars and specified colors
ggplot(filtered_data, aes(x = as.factor(roundNum), fill = as.factor(investorState))) +
  geom_bar(aes(group = investorState), position = position_stack(reverse = FALSE), stat = "count") +
  scale_fill_manual(values = state_colors) +
  facet_wrap(~condition.f*gameNum.f, scales = "free_x", labeller = label_both) +
  labs(x = "Round", fill = "Investor State", title = "Histogram of Latent States per Game and Round") +
  theme_bw() +
  theme(legend.position = "bottom")

```{r}
counts <- filtered_data %>%
  group_by(condition.f, gameNum.f, investorState,id) %>%
  summarise(count = n()) %>%
    summarise(count = sum(count))

# Calculate the average count per state per condition
total_rounds <- filtered_data  %>%
  group_by(condition.f, gameNum.f) %>%
  summarise(total_rounds = n())

@@

ggplot(state_summary , aes(x = gameNum.f, y = state_percentage, fill = as.factor(investorState))) +
  geom_bar(stat = "identity") +
  facet_wrap(~condition.f) +
  labs(x = "Condition", y = "Proportion", fill = "Investor State") +
  theme_bw() +
  theme(legend.position = "bottom")

@@

# Create the histogram with stacked bars and specified colors
ggplot(state_summary , aes(x = gameNum.f, y = state_percentage, fill = as.factor(investorState))) +
  geom_bar(aes(group = investorState), position = position_stack(reverse = FALSE), stat = "identity") +
  scale_fill_manual(values = state_colors) +
  facet_wrap(~condition.f*gameNum.f, scales = "free_x", labeller = label_both) +
  labs(x = "Round", fill = "Investor State", title = "Histogram of Latent States per Game and Round") +
  theme_bw() +
  theme(legend.position = "bottom")
```

>>>>

# Find IDs for those who participated in experiment and add RS_score and high_RS as new columns to d_finished.
# Merging d_finished with relevant columns from RS_data based on matching ids
d_finished <- d_finished %>%
  inner_join(RS_data %>% dplyr::select(id, RS_score, high_RS), by = "id")

@@

# View the summarized counts
print(temp)

###################### Extract latent Investor State in each round ####################

# Select the desired columns
d_selected <- d_finished %>%
  dplyr::select(roundNum, gameNumber, playerId, starts_with("State_"))

# Reshape the data
d_reshaped <- d_selected %>%
  pivot_longer(cols = starts_with("State_"), 
               names_to = "State_name", 
               values_to = "investorState") %>%
  mutate(roundNum = as.numeric(str_extract(State_name, "(?<=r)\\d+")),
         gameNumber = as.numeric(str_extract(State_name, "(?<=game_)\\d+"))) %>%
  dplyr::select(playerId, gameNumber, roundNum, investorState) %>%
  arrange(playerId, gameNumber, roundNum) %>%
  filter(!is.na(investorState)) %>% unique()

# Join the reshaped data back to the original dataset
d_finished <- d_finished %>%
  dplyr::select(-investorState) %>% 
  left_join(d_reshaped, by = c("playerId", "gameNumber", "roundNum"))
```

@@

d_finished <- d_finished  %>% mutate( gameNum.f = factor(gameNumber,labels = c("pre", "expo1", "expo2", "expo3","post"),
                                      levels=c("1", "11", "12", "13", "2")),
                                      phase.f = factor(ifelse(gameNumber %in% c("1", "2"), 
                                                       ifelse(gameNumber == "1", "pre", "post"), NA), levels = c("pre", "post")),
                                      investorState.f = factor(investorState, labels= c("unhappy", "neutral", "happy"), levels =c("unhappy", "neutral", "happy"))) %>% 
                              rename(# Before Exposure
                            rating_cooperative_1 = `ratePLayer_trustGame_1.How cooperative do you think the other player is?`,
                            rating_forgiving_1 = `ratePLayer_trustGame_1.How forgiving do you think the other player is?`,
                            rating_playAgain_1 = `ratePLayer_trustGame_1.If given the choice, would you like to play again with the other player?`,

@@

                            )

@@

# Calculate means using vectorized operations
d_finished <- d_finished %>%
  mutate(
    LPFS_score = rowMeans(dplyr::select(., all_of(lpfs_cols)), na.rm = TRUE),
    LPFS_self_score = rowMeans(dplyr::select(., all_of(lpfs_1_to_6_cols)), na.rm = TRUE),
    LPFS_other_score = rowMeans(dplyr::select(., all_of(lpfs_7_to_12_cols)), na.rm = TRUE)
  )
save(d_finished, file = "data/d_finished.RData")

@@

avg_ret_df <- d_finished %>%
  dplyr::select(playerId,roundType,investment,returns,roundNum,gameNum.f,condition.f, phase.f, high_RS, LPFS_score, RS_score, investorState.f) %>% 
  filter(roundType=="trust",!is.na(roundNum)) %>% 
  mutate(roundNum = as.numeric(as.character(roundNum))) %>%
  mutate(inv_scaled = as.vector(scale(investment))) %>% 
  mutate(inv_pct = investment/20, ret_pct_0 = ifelse(investment==0,0,returns/(3*investment)),ret_pct_na = ifelse(investment==0,NA,returns/(3*investment)))

@@

```{r modAllReturns, include=FALSE, cache=TRUE}
##############  Filter only pre and post games
prepost_df <- avg_ret_df %>% filter(gameNum.f %in% c("pre","post"))
##############
mod_returns_pct_RS <- mixed( ret_pct_na ~ phase.f*condition.f*inv_scaled*high_RS + (1 + phase.f| playerId), prepost_df, REML= TRUE, method="KR")

@@

```{r}
contrasts(prepost_df$phase.f)
contrasts(prepost_df$condition.f)
contrasts(prepost_df$high_RS)
contrasts(prepost_df$investorState.f)
```

@@

# Compute EMMs for the interaction of interest
emms <- emmeans(mod_returns_pct_RS, ~ phase.f * condition.f | high_RS, pbkrtest.limit = 6300, adjust="tukey")

@@

<!---------------- TRYING A MODEL WITH LATENT HMM STATE AS COVARIATE INSTEAD OF INVESTMENT  ------------------->

```{r}
mod_returns_latent <- mixed( ret_pct_na ~ phase.f*condition.f*investorState.f*high_RS + (1 + phase.f| playerId), prepost_df, REML= TRUE, method="KR")

summary(mod_returns_latent)
```

```{r, include= F}
# Load necessary packages
library(lme4)
library(emmeans)
library(ggplot2)
library(dplyr)

# Assuming `mod_returns_latent` is your fitted model and `prepost_df` is your data frame
# Calculate estimated marginal means
emm <- emmeans(mod_returns_latent, ~ high_RS*condition.f*phase.f   | investorState.f, adjust = "tukey")

# Calculate pairwise comparisons
contrast_results <- contrast(emm, method = "pairwise")

# Extract significant contrasts
significant_contrasts <- summary(contrast_results, infer = c(TRUE, TRUE))

# Convert emmeans object to data frame for plotting
emm_df <- as.data.frame(emm)

# Plot the marginal means using ggplot2
p <- ggplot(emm_df, aes(x = investorState.f, y = emmean, color = high_RS, group = high_RS)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2, position = position_dodge(width = 0.5)) +
  geom_line(position = position_dodge(width = 0.5), aes(linetype = high_RS)) +
  labs(title = "Marginal Means of Returns by Investor State and RS Group",
       x = "Investor State",
       y = "Estimated Marginal Mean of Returns") +
  theme_minimal() +
  scale_color_manual(name = "RS Group", values = c("low RS" = "blue", "high RS" = "red"))

# Add significance annotations
for(i in 1:nrow(significant_contrasts)) {
  row <- significant_contrasts[i, ]
  if(row$p.value < 0.05) {
    p <- p + annotate("text", 
                      label = "*", 
                      x = which(levels(emm_df$investorState.f) == row$investorState.f),
                      y = max(emm_df$emmean) + 0.05 * i,
                      color = "black", 
                      size = 5, 
                      vjust = -0.5)
  }
}

# Print the plot
print(p)

```
```{r}
# Load necessary packages
library(lme4)
library(emmeans)
library(ggplot2)
library(dplyr)
library(ggsignif)

# Assuming `mod_returns_latent` is your fitted model and `prepost_df` is your data frame
# Calculate estimated marginal means
emm <- emmeans(mod_returns_latent, ~  high_RS | investorState.f  )

# Calculate pairwise comparisons
contrast_results <- pairs(emm)

print(contrast_results)

# Convert emmeans object to data frame for plotting
emm_df <- as.data.frame(emm)

# Convert contrast results to data frame for annotation
contrast_df <- as.data.frame(contrast_results)
significant_contrasts <- contrast_df %>%
  filter(p.value < 0.05) %>%
  mutate(significance = ifelse(p.value < 0.001, "***", ifelse(p.value < 0.01, "**", "*")))

print(significant_contrasts)

# Plot the marginal means using ggplot2
p <- ggplot(emm_df, aes(x = investorState.f, y = emmean, color = high_RS, group = high_RS)) +
  geom_point(position = position_dodge(width = 0.8), size = 3) +
  geom_errorbar(aes(ymin = emmean - SE, ymax = emmean + SE), width = 0.2, position = position_dodge(width = 0.8)) +
  geom_line(position = position_dodge(width = 0.8), aes(linetype = high_RS)) +
  labs(title = "Marginal Means of Returns by Investor State and RS Group",
       x = "Investor State",
       y = "Estimated Marginal Mean of Returns") +
  theme_minimal() +
  scale_color_manual(name = "RS Group", values = c("low RS" = "blue", "high RS" = "red")) +
  theme(plot.title = element_text(size = 16), legend.position = "bottom")

# Print the plot
print(p)

@@

# Generate the plot with points and error bars
ret_all <- afex_plot(mod_returns_pct_RS, x = "condition.f", trace = "phase.f", dodge = 0.8, error = "within",
                     mapping = c("linetype", "shape", "fill"),
                     factor_levels = list(condition.f = c("Control", "Manipulation")),
                     legend_title = "Phase",
                     emmeans_arg = list(p.adjust.method = "tukey"))

@@

# Convert dataset from wide to long format, focusing on columns that start with "State_"
d_long <- d_finished %>%
  dplyr::select(id, condition.f, gameNum.f, gameNumber, roundNum, investorState.f) %>% 
  drop_na()
# View the resulting dataset
print(d_long)

@@

# Filter for only games 1 and 2
filtered_prepost <- d_long %>%
  filter(gameNum.f %in% c("pre", "post"))

@@

# Create the histogram with stacked bars and specified colors
ggplot(filtered_prepost, aes(x = as.factor(roundNum), fill = investorState.f)) +
  geom_bar(aes(group = investorState.f), position = position_stack(reverse = FALSE), stat = "count") +
  scale_fill_manual(values = state_colors) +
  facet_wrap(~condition.f*gameNum.f, scales = "free_x", labeller = label_both) +
  labs(x = "Round", fill = "Investor State", title = "Histogram of Latent HMM Investor States per Game and Round") +
  theme_bw() +
  theme(legend.position = "bottom")

```
```{r}
# Filter for expo games only
filtered_expo <- d_long %>%
  filter(gameNum.f %in% c("expo1", "expo2", "expo3"))

# Create the histogram with stacked bars and specified colors
ggplot(filtered_expo, aes(x = as.factor(roundNum), fill = investorState.f)) +
  geom_bar(aes(group = investorState.f), position = position_stack(reverse = FALSE), stat = "count") +
  scale_fill_manual(values = state_colors) +
  facet_wrap(~condition.f*gameNum.f, scales = "free_x", labeller = label_both) +
  labs(x = "Round", fill = "Investor State", title = "Histogram of Latent States per Game and Round") +
  theme_bw() +
  theme(legend.position = "bottom")
```

<!------------------------ AVERAGING LATENT STATE COUNT ACROSS TRIALS ----------------------------->

```{r}
counts <- filtered_prepost %>%
  group_by(condition.f, gameNum.f, investorState.f,id) %>%
  summarise(count = n()) %>%
    summarise(count = sum(count))

# Calculate the average count per state per condition
total_rounds <- filtered_prepost  %>%
  group_by(condition.f, gameNum.f) %>%
  summarise(total_rounds = n())

@@

ggplot(state_summary , aes(x = gameNum.f, y = state_percentage, fill = investorState.f)) +
  geom_bar(stat = "identity") +
  facet_wrap(~condition.f) +
  labs(x = "Condition", y = "Proportion", fill = "Investor State") +
  theme_bw() +
  theme(legend.position = "bottom")

@@

# Create the histogram with stacked bars and specified colors
ggplot(state_summary , aes(x = gameNum.f, y = state_percentage, fill = investorState.f)) +
  geom_bar(aes(group = investorState.f), position = position_stack(reverse = FALSE), stat = "identity") +
  scale_fill_manual(values = state_colors) +
  facet_wrap(~condition.f*gameNum.f, scales = "free_x", labeller = label_both) +
  labs(x = "Round", fill = "Investor State", title = "Histogram of Latent States per Game and Round") +
  theme_bw() +
  theme(legend.position = "bottom")
```
